---
title: "Project Proposal"
author: "Karlmarx Odhiambo"
date: "`r Sys.Date()`"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Using Logistic Regression to Predict Heart Disease

The main aim of this project is to develop a logistic regression model that can be able to predict the posibility of heart disease to an individual. According to the WHO, 12 million people die annually from Heart-related conditions. Furthermore, over half of the total deaths in America are due to cardiovascular diseases. An early identification of such conditions can help patients make important decisions on lifestyle changes to improve their health status. I intend to use this project to highlight the risk factors of heart disease and predict the overall risk using logistic regression. 

## Data Explanation

I sourced this data from the Kaggle Website. The data was collected by the World Health Organization from the residents of Framingham, Massachusetts. It has over 4000 subjects and 16 variables.

# Data Dictionary

The data has both demographic, behavioral and medical risk factors.

**Demographic:**
• Sex: male or female(Nominal)
• Age: Age of the patient;(Continuous - Although the recorded ages have been truncated to whole numbers, the concept of age is continuous)
**Behavioral:**
• Current Smoker: whether or not the patient is a current smoker (Nominal)
• Cigs Per Day: the number of cigarettes that the person smoked on average in one day.(can be considered continuous as one can have any number of cigarettes, even half a cigarette.)
**Medical( history):**
• BP Meds: whether or not the patient was on blood pressure medication (Nominal)
• Prevalent Stroke: whether or not the patient had previously had a stroke (Nominal)
• Prevalent Hyp: whether or not the patient was hypertensive (Nominal)
• Diabetes: whether or not the patient had diabetes (Nominal)
**Medical(current):**
• Tot Chol: total cholesterol level (Continuous)
• Sys BP: systolic blood pressure (Continuous)
• Dia BP: diastolic blood pressure (Continuous)
• BMI: Body Mass Index (Continuous)
• Heart Rate: heart rate (Continuous - In medical research, variables such as heart rate though in fact discrete, yet are considered continuous because of large number of possible values.)
• Glucose: glucose level (Continuous)
**Predict variable (desired target):**
• 10 year risk of coronary heart disease CHD (binary: “1”, means “Yes”, “0” means “No”)

First, let's load some libraries to be used in the project

```{r}
library(tidyverse)
library(dplyr)
library(cowplot)
library(pROC)
```


Let me import the data to highlight some of the attributes of the data

```{r}
framingham <- read.csv("C:/Users/karlm/OneDrive/Documents/STA 631/framingham.csv")
```

Below is an overview of how I will handle my project

## Data Exploration

```{r}
# View the structure of the dataset
str(framingham)

# Summary statistics
summary(framingham)

# Check for missing values
colSums(is.na(framingham))

# Check the class distribution of the target variable
table(framingham$TenYearCHD)
```

```{r}
head(df)
```

```{r}
cat('Number of Rows:',nrow(df),'   ','Number of Fields:',ncol(df))
```

### EDA

## Handle missing values

Depending on the extent of missing data, I will impute missing values or remove rows with missing values.

## Convert categorical variables to factors

```{r}
framingham$education <- as.factor(framingham$education)
framingham$BPMeds <- as.factor(framingham$BPMeds)
# Similarly, convert other categorical variables to factors if needed
```

## Split the data into training and testing sets

```{r split data}
# set.seed(123) # for reproducibility
# train_index <- createDataPartition(framingham$TenYearCHD, p = 0.7, list = FALSE)
# train_data <- framingham[train_index, ]
# test_data <- framingham[-train_index, ]
```


### Build and Evaluate the Logistic Regression Model

**Build the logistic regression model**

```{r lrm}
# model <- glm(TenYearCHD ~ ., data = train_data, family = binomial)
```


**Summary of the model**

```{r}
# summary(model)
```


**Make predictions on the test data**

```{r}
# predictions <- predict(model, newdata = test_data, type = "response")
```


**Convert predicted probabilities to binary predictions**

```{r}
# predicted_classes <- ifelse(predictions > 0.5, 1, 0)
```


**Confusion matrix**

```{r}
# confusionMatrix(table(predicted_classes, test_data$TenYearCHD))
```


## Interpretation and Further Analysis

*Further steps*
Interpret the coefficients of the model to understand the impact of each predictor on the probability of TenYearCHD.

Evaluate the model's performance using metrics like ROC curve, AUC, etc using the pROC library.

Consider additional analyses:

Feature selection or variable importance analysis to identify the most important predictors.

Assess model assumptions such as linearity and independence of errors.

Communicate results:

Present findings in a clear and understandable manner









